---
title: "Analysis of USA Storm Data from 1950 - 2011"
author: "Yolande Korsten"
date: "Sunday, August 24, 2014"
output: html_document
---

The purpose of this analysis of the NOAA storm data from 1950 - 2011 is to identify which types of events are most harmful in respect to public health over time, and also which events have the greatest economic conseqences. 

## Data Processing

```{r echo=TRUE, cache=TRUE, results='hide'}
rawStorm <- read.csv(bzfile("./repdata-data-StormData.csv.bz2"))
```

The date data is stored as a factor variable.  Converting it to class Date will make analysis later easier.

```{r echoe=TRUE, cache=TRUE}
class(rawStorm$BGN_DATE)
```

```{r echoe=TRUE, cache=TRUE}
class(rawStorm$END_DATE)
```

```{r echo=TRUE, cache=TRUE}
## Converting Date Fields from class factor to class Date

rawStorm$BGN_DATE<-as.character(rawStorm$BGN_DATE)
rawStorm$BGN_DATE <- as.Date(rawStorm$BGN_DATE, format="%m/%d/%Y")
rawStorm$END_DATE<-as.character(rawStorm$END_DATE)
rawStorm$END_DATE <- as.Date(rawStorm$END_DATE, format="%m/%d/%Y")

```

```{r echoe=TRUE, cache=TRUE}
class(rawStorm$BGN_DATE)
```

```{r echoe=TRUE, cache=TRUE}
class(rawStorm$END_DATE)
```

There are 2 fields with economic data which indicate the damage caused in a non-numeric format eg. 2.5 M means 2.5 Million. These descriptive terms need to be converted into the actual numeric values e.g. 25 000 000.

```{r echoe=TRUE, cache=TRUE}
print(summary(rawStorm[,c("PROPDMG","CROPDMG")]))
```

```{r echo=TRUE, cache=TRUE}
#convert accepts a data.frame with 2 columns
#x[1] - this should be a value
#x[2] = this is a indicator of K, M, B to indicate the EXP
# Returns the actual value
convert<-function(x) {
  value=0;
  if (!is.na(x[2])){
    if (x[2]=="K") {
      value=as.numeric(x[1])*1000
    }
    if (x[2]=="M") {
      value=as.numeric(x[1])*1000000
    }
    if (x[2]=="B") {
      value=as.numeric(x[1])*1000000000
    }
  }
  return (value)
} 

## Converting the economic data into actual values

rawStorm$PROPDMG <- apply(rawStorm[ , c("PROPDMG","PROPDMGEXP")], 1, FUN=convert)
rawStorm$CROPDMG <- apply(rawStorm[ , c("CROPDMG","CROPDMGEXP")], 1, FUN=convert)

print(summary(rawStorm[,c("PROPDMG","CROPDMG")]))
```





## Analysis

The following assumptions regarding the data has been made:  
* Public Health impact is measured by injuries (INJURIES) or fatalities (FATALITIES) during the event.  
* Economic consequence is measured by crop damage (CROPDMG) and property damage (PROPDMG). 

### Health Impact

**Evaluating the most destructive events on public health by Type.**

```{r echo=TRUE, cache=TRUE}
library(plyr)
## Calculating aggregate injuries by Event Type

evInjuries <- aggregate(rawStorm$INJURIES, by = list(rawStorm$EVTYPE), FUN = sum)
evFatal <- aggregate(rawStorm$FATALITIES, by = list(rawStorm$EVTYPE), FUN = sum)
pHealthEV <- cbind(evInjuries, evFatal[,2])
colnames(pHealthEV) <- c("Event", "totInjuries", "totFatal")
pHealthEV <- mutate(pHealthEV, totOuch=apply(pHealthEV[,2:3], 1, sum))
head(pHealthEV)

```

```{r echo=TRUE, cache=TRUE}
## Reordering the information to be in descending order
opHealthEV <- pHealthEV[order(pHealthEV$totOuch,decreasing=TRUE), ]
spHealthEV <- opHealthEV[1:15,]

spHealthEV$Event <- factor(spHealthEV$Event)

head(spHealthEV)
```




```{r echo=FALSE, fig.width=10,  fig.height=5}
library(ggplot2)
# Map the time of day to different fill colors. These both have the same result.
gpHealthEVs <- ggplot(data=spHealthEV, aes(x=Event, y=totOuch, fill=Event)) + 
      geom_bar(stat="identity", horiz=TRUE) + 
      coord_flip() + 
      ggtitle("Public Health Impact by Event") + 
      xlab("Event Type") +
      ylab("Total injuries & fatalities")

gpHealthEVs
```


### Economic Impact

**Evaluating the most economically destructive events by TYPE**

```{r echo=TRUE, cache=TRUE}
library(plyr)
## Calculating aggregate economic damage by Event Type

evProp <- aggregate(rawStorm$PROPDMG, by = list(rawStorm$EVTYPE), FUN = sum)
evCrop <- aggregate(rawStorm$CROPDMG, by = list(rawStorm$EVTYPE), FUN = sum)
evHealth <- cbind(evProp, evCrop[,2])
colnames(evHealth) <- c("Event", "totProp", "totCrop")
evHealth <- mutate(evHealth, totEOuch=apply(evHealth[,2:3], 1, sum))
head(evHealth)

```

```{r echo=TRUE, cache=TRUE}
## Reordering the information to be in descending order
oevHealth <- evHealth[order(evHealth$totEOuch,decreasing=TRUE), ]
sevHealth <- oevHealth[1:15,]

sevHealth$Event <- factor(sevHealth$Event)

head(sevHealth)
```


```{r echo=FALSE, fig.width=10,  fig.height=5}
library(ggplot2)

gevHealths <- ggplot(data=sevHealth, aes(x=Event, y=totEOuch, fill=Event)) + 
      geom_bar(stat="identity", horiz=TRUE) + 
      coord_flip() + 
      ggtitle("Economic Destruction by Event") + 
      xlab("Event Type") +
      ylab("Total Value of Damage to Crops and Property")

gevHealths

```

## Results

The results of the analysis shows that while there is a large variability in terms of the day that the most damage occurs for either economic or public health, the following Events have proven to be the most destructive.

**Public Health**
  
1. Tornados  
2. Excessive Heat    
3. Floods  
4. Thunderstorm Winds  
5. Lightning  
  
**Economic**  

1. Floods  
2. Hurricanes/Typhoons  
3. Tornados  
4. Storm Surges  
5. Hail 
  

